---
title: "RforDataScience_Wrangle"
author: "Victoria"
date: "2022-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r}
library(tidyverse)
library(nycflights13)
library(gapminder)
library(Lahman)
```

# Wrangle
## Tibbles
Load tidyverse
```{r}
library(tidyverse) # mainly using the tibble package
```

### Creating tibbles
Coerce a dataframe to a tibble using as_tibble():
```{r}
as_tibble(iris)
```

Create a new tibble from individual vectors using tibble():
```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

Can include column name sin tibbles that aren't valid R variable names by surrounding them in backticks
```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

Create a transposed tibble with tribble():
```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

### Tibbles vs Dataframe
#### Printing
Printing a tibble will only show the first 10 rows and all cols that will fit on the screen.
Each col will report the data type it contains (like str()):
```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

Options to print more than the default output:
First, you can explicitly print() the data frame and control the number of rows (n) and the width of the display. width = Inf will display all columns:
```{r}
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

You can also control the default print behaviour by setting options:
- options(tibble.print_max = n, tibble.print_min = m): if more than n rows, print only m rows. Use options(tibble.print_min = Inf) to always show all rows.
- Use options(tibble.width = Inf) to always print all columns, regardless of the width of the screen.

You can see a complete list of options by looking at the package help with package?tibble.

Can also view the entire dataset using R's buit-in data viewer, View()
```{r}
nycflights13::flights %>% 
  View()
```

#### Subsetting
 If you want to pull out a single variable, you need some new tools, $ and [[]]. [[]] can extract by name or position; $ only extracts by name but is a little less typing.
 
```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161

# Extract by position
df[[1]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
```
 
To use these in a pipe, you'll need to use "." as a placeholder:
```{r}
df %>% .$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df %>% .[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
```

Some older functions won't work on tibbles. In that case, convert the tibble to a data frame with as.data.frame().

## Data import
Load packages
```{r}
library(tidyverse) # will be using the readr package
```

### readr functions
read_csv() reads comma delimited files, read_csv2() reads semicolon separated files (common in countries where , is used as the decimal place), read_tsv() reads tab delimited files, and read_delim() reads in files with any delimiter.

read_fwf() reads fixed width files. You can specify fields either by their widths with fwf_widths() or their position with fwf_positions(). read_table() reads a common variation of fixed width files where columns are separated by white space.

read_log() reads Apache style log files. (But also check out webreadr which is built on top of read_log() and provides many more helpful tools.)

### Parsing a vector
parse_*() functions take a character vector and return a more specialised vector like a logical, integer, or date:
```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
#>  logi [1:3] TRUE FALSE NA
str(parse_integer(c("1", "2", "3")))
#>  int [1:3] 1 2 3
str(parse_date(c("2010-01-01", "1979-10-14")))
#>  Date[1:2], format: "2010-01-01" "1979-10-14"
```

Using parsers is mostly a matter of understanding what’s available and how they deal with different types of input. There are eight particularly important parsers:

parse_logical() and parse_integer() parse logicals and integers respectively. There’s basically nothing that can go wrong with these parsers.

parse_double() is a strict numeric parser, and parse_number() is a flexible numeric parser. These are more complicated than you might expect because different parts of the world write numbers in different ways.

parse_character() seems so simple that it shouldn’t be necessary. But one complication makes it quite important: character encodings.

parse_factor() create factors, the data structure that R uses to represent categorical variables with fixed and known values.

parse_datetime(), parse_date(), and parse_time() allow you to parse various date & time specifications. These are the most complicated because there are so many different ways of writing dates.

### Parsing a file
Guess the data type of each column using guess_parser() and parse_guess()
guess_parser() returns readr's best guess, and
parse_guess() uses that guess to parse the column
```{r}
guess_parser("2010-10-01")
#> [1] "date"
guess_parser("15:01")
#> [1] "time"
guess_parser(c("TRUE", "FALSE"))
#> [1] "logical"
guess_parser(c("1", "5", "9"))
#> [1] "double"
guess_parser(c("12,352,561"))
#> [1] "number"

str(parse_guess("2010-10-10"))
#>  Date[1:1], format: "2010-10-10"
```

### Writing to a file
```{r}
# write_csv()
write_csv(challenge, "challenge.csv")
```

Can also use write_tsv() or write_RDS()

## Tidy data
```{r}
library(tidyverse) # using the tidyr package this chapter
```

There are three interrelated rules which make a dataset tidy:
- Each variable must have its own column.
- Each observation must have its own row.
- Each value must have its own cell.

That interrelationship leads to an even simpler set of practical instructions:
- Put each dataset in a tibble.
- Put each variable in a column.

### Pivoting
Pivot (with pivot_longer() or pivot_wider()) to resolve one of two common problems:
One variable might be spread across multiple columns.
One observation might be scattered across multiple rows.

#### pivot_longer()
Column names may not be names of variables, but values of a variable.
Pivot the offending columns into a new variables. Will need 3 parameters:
- The set of columns whose names are values, not variables.
- The name of the variable to move the column names to.
- The name of the variable to move the column values to.

call pivot_longer() with these parameters:
```{r}
table4a # cols 1999 and 2000 are both years

table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

Note that pivot_longer() makes datasets longer by increasing the nrows but decreasing ncol.

Tidy table4b similarly
```{r}
table4b # same issue with years in different cols

table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

Combined tidies versions of table4a/b with dplyr::left_join()
```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

left_join(tidy4a, tidy4b)
```

#### pivot_wider()
Use pivot_wider when an observation is scattered across multiple rows. Need 2 parameters:
- The column to take variable names from
- The column to take values from

```{r}
table2 # observation = country in a year, but this is spread across rows

table2 %>%
    pivot_wider(names_from = type, values_from = count)
```

Note that pivot_wider() does the opposite of pivot_longer() - it decreases the nrow and increases ncol

### Separating and uniting
#### separate()
separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears
```{r}
table3 # rate col contains both cases and population variables

table3 %>% 
  separate(rate, into = c("cases", "population"))
```

```{r}
# convert cases and population vectors to integer type
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

```{r}
# split year into century and year - set up for next section
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

#### unite()
unite() combines multiple columns into a single column

```{r}
# rejoin the century and year cols into one year col
table5 %>% 
  unite(new, century, year)

# remove the default _ separator
table5 %>% 
  unite(new, century, year, sep = "")
```

### Missing values
Values can be missing in 2 ways:
- Explicitly, i.e. flagged with NA.
- Implicitly, i.e. simply not present in the data.

Example dataset
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

There are two missing values in this dataset:
- The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains NA.
- The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

Can make an implicit value explicit by putting years in the columns:
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```

Then set values_drop_na = TRUE to turn explicit missing values implicit
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

Can also use complete() to make missing values explicit in tidy data
```{r}
stocks %>% 
  complete(year, qtr)
```

 Sometimes when a data source has primarily been used for data entry, missing values indicate that the previous value should be carried forward:
```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
```

Can fill in these missing values with fill()
```{r}
treatment %>% 
  fill(person)
```

