---
title: "RforDataScience_Wrangle"
author: "Victoria"
date: "2022-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r}
library(tidyverse)
library(nycflights13)
library(gapminder)
library(Lahman)
```

# Wrangle
## Tibbles
Load tidyverse
```{r}
library(tidyverse) # mainly using the tibble package
```

### Creating tibbles
Coerce a dataframe to a tibble using as_tibble():
```{r}
as_tibble(iris)
```

Create a new tibble from individual vectors using tibble():
```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

Can include column name sin tibbles that aren't valid R variable names by surrounding them in backticks
```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

Create a transposed tibble with tribble():
```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

### Tibbles vs Dataframe
#### Printing
Printing a tibble will only show the first 10 rows and all cols that will fit on the screen.
Each col will report the data type it contains (like str()):
```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

Options to print more than the default output:
First, you can explicitly print() the data frame and control the number of rows (n) and the width of the display. width = Inf will display all columns:
```{r}
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

You can also control the default print behaviour by setting options:
- options(tibble.print_max = n, tibble.print_min = m): if more than n rows, print only m rows. Use options(tibble.print_min = Inf) to always show all rows.
- Use options(tibble.width = Inf) to always print all columns, regardless of the width of the screen.

You can see a complete list of options by looking at the package help with package?tibble.

Can also view the entire dataset using R's buit-in data viewer, View()
```{r}
nycflights13::flights %>% 
  View()
```

#### Subsetting
 If you want to pull out a single variable, you need some new tools, $ and [[]]. [[]] can extract by name or position; $ only extracts by name but is a little less typing.
 
```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161

# Extract by position
df[[1]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
```
 
To use these in a pipe, you'll need to use "." as a placeholder:
```{r}
df %>% .$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df %>% .[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
```

Some older functions won't work on tibbles. In that case, convert the tibble to a data frame with as.data.frame().

## Data import
Load packages
```{r}
library(tidyverse) # will be using the readr package
```

### readr functions
read_csv() reads comma delimited files, read_csv2() reads semicolon separated files (common in countries where , is used as the decimal place), read_tsv() reads tab delimited files, and read_delim() reads in files with any delimiter.

read_fwf() reads fixed width files. You can specify fields either by their widths with fwf_widths() or their position with fwf_positions(). read_table() reads a common variation of fixed width files where columns are separated by white space.

read_log() reads Apache style log files. (But also check out webreadr which is built on top of read_log() and provides many more helpful tools.)

### Parsing a vector
parse_*() functions take a character vector and return a more specialised vector like a logical, integer, or date:
```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
#>  logi [1:3] TRUE FALSE NA
str(parse_integer(c("1", "2", "3")))
#>  int [1:3] 1 2 3
str(parse_date(c("2010-01-01", "1979-10-14")))
#>  Date[1:2], format: "2010-01-01" "1979-10-14"
```

Using parsers is mostly a matter of understanding what’s available and how they deal with different types of input. There are eight particularly important parsers:

parse_logical() and parse_integer() parse logicals and integers respectively. There’s basically nothing that can go wrong with these parsers.

parse_double() is a strict numeric parser, and parse_number() is a flexible numeric parser. These are more complicated than you might expect because different parts of the world write numbers in different ways.

parse_character() seems so simple that it shouldn’t be necessary. But one complication makes it quite important: character encodings.

parse_factor() create factors, the data structure that R uses to represent categorical variables with fixed and known values.

parse_datetime(), parse_date(), and parse_time() allow you to parse various date & time specifications. These are the most complicated because there are so many different ways of writing dates.

### Parsing a file
Guess the data type of each column using guess_parser() and parse_guess()
guess_parser() returns readr's best guess, and
parse_guess() uses that guess to parse the column
```{r}
guess_parser("2010-10-01")
#> [1] "date"
guess_parser("15:01")
#> [1] "time"
guess_parser(c("TRUE", "FALSE"))
#> [1] "logical"
guess_parser(c("1", "5", "9"))
#> [1] "double"
guess_parser(c("12,352,561"))
#> [1] "number"

str(parse_guess("2010-10-10"))
#>  Date[1:1], format: "2010-10-10"
```

### Writing to a file
```{r}
# write_csv()
write_csv(challenge, "challenge.csv")
```

Can also use write_tsv() or write_RDS()


