---
title: "tidyverse_packages"
author: "Victoria"
date: "2022-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages
```{r}
# install.packages("tidyverse")
# install.packages(c("nycflights13", "gapminder", "Lahman"))
```

# Load packages
```{r}
library(tidyverse)
library(nycflights13)
library(gapminder)
library(Lahman)
```

Introduction to tidyverse is given in the R for Data Science book. Online pdf is given here: https://r4ds.had.co.nz/

# Explore
## Data visualization

Load the ggplot2 mpg data frame
```{r}
library(ggplot2)
mpg
```

### Creating a ggplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

A graphing template
```{r}
# ggplot(data = <DATA>) + 
 # <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

Exercise
```{r}
ggplot(data = mpg)
```

```{r}
nrow(mpg)
ncol(mpg)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv)) # doesn't give number of vehicles under each category
```

### Aesthetic mapping
You can add a third variable, like class, to a two dimensional scatterplot by mapping it to an aesthetic.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

### Common problems
```{r}
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) # + has to be at the end of a line, not the start
```

### Facets
Facets are subplots that each display one subset of the data.

#### facet_wrap()
```{r}
# facet by a single variable 
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

#### facet_grid()
```{r}
# facet by multiple variables
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

### Geometric objects
A geom is the geometrical object that a plot uses to represent data. People often describe plots by the type of geom that the plot uses. For example, bar charts use bar geoms, line charts use line geoms, boxplots use boxplot geoms, and so on. Scatterplots break the trend; they use the point geom.

To change the geom in your plot, change the geom function that you add to ggplot()

```{r}
# left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# right
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

To display multiple geoms in the same plot, add multiple geom functions to ggplot()

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

You can use the same idea to specify different data for each layer. Here, our smooth line displays just a subset of the mpg dataset, the subcompact cars. The local data argument in geom_smooth() overrides the global data argument in ggplot() for that layer only.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

### Statistical transformations
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

Can generally use geom and stats interchangeably - bc every stat has a default geom.

```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

### Position adjustments
Can color a plot with color or fill

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

Mapping the fill aesthetic to another variable automatically stacks the bars - due to position adjustment.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

 If you don’t want a stacked bar chart, you can use one of three other options: "identity", "dodge" or "fill"
 
#### identity
position = "identity" will place each object exactly where it falls in the context of the graph. This is not very useful for bars, because it overlaps them. To see that overlapping we either need to make the bars slightly transparent by setting alpha to a small value, or completely transparent by setting fill = NA.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
 
#### dodge
position = "fill" works like stacking, but makes each set of stacked bars the same height. This makes it easier to compare proportions across groups.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

#### fill
position = "dodge" places overlapping objects directly beside one another. This makes it easier to compare individual values.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

#### jitter
position = "jitter" adds a small amount of random noise to each point. This spreads the points out because no two points are likely to receive the same amount of random noise.
*specifically for scatterplots

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

### Coordinate systems
#### cord_flip()
coord_flip() switches the x and y axes. This is useful (for example), if you want horizontal boxplots. It’s also useful for long labels: it’s hard to get them to fit without overlapping on the x-axis.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

#### cord_quickmap()
coord_quickmap() sets the aspect ratio correctly for maps. This is very important if you’re plotting spatial data with ggplot2

```{r}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

#### cord_polar()
coord_polar() uses polar coordinates. Polar coordinates reveal an interesting connection between a bar chart and a Coxcomb chart

```{r}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

### Layered grammar of graphics
Code template:
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
  
  *Rarely need to supply all parameters (bracketed above), since ggplot2 provides defaults for all except data, mappings, and the geom function.
  
## Data Transformation
How to transform data using the dplyr package and a dataset on flights departing NYC in 2013. 

### Load packages
```{r}
library(nycflights13)
library(tidyverse) # will use both dplyr & ggplot2
```

### nycflights13 data
```{r}
flights # note that this is a tibble, a df unique to the tidyverse
```

### dplyr basics
The dplyr package contains the following functions:
- Pick observations by their values (filter()).
- Reorder the rows (arrange()).
- Pick variables by their names (select()).
- Create new variables with functions of existing variables (mutate()).
- Collapse many values down to a single summary (summarise()).

All verbs work similarly:
1. The first argument is a data frame.
2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).
3. The result is a new data frame.

### filter()
Use to subset select rows
```{r}
# select all flights on Jan 1st
filter(flights, month == 1, day == 1)
```
Can both save results in a variable and still print the result in one line by wrapping the assignment in ()
```{r}
(dec25 <- filter(flights, month == 12, day == 25))
```

filter() only includes rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:
```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)

# vs

filter(df, is.na(x) | x > 1)
```

### arrange()
Use to change the order of rows
```{r}
arrange(flights, year, month, day)
```

Use desc to re-order by a column in descending order
```{r}
arrange(flights, desc(dep_delay))
```

Missing values are always sorted at the end
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

### select()
Allows you to rapidly zoom in on a useful subset of a large dataset using operations based on the names of variables.

```{r}
# select columns by name
select(flights, year, month, day)
```

```{r}
# select columns between year and day 
select(flights, year:day)
```

```{r}
# select all columns except those from year to day
select(flights, -(year:day))
```

There are a number of helper functions you can use within select():
- starts_with("abc"): matches names that begin with “abc”.
- ends_with("xyz"): matches names that end with “xyz”.
- contains("ijk"): matches names that contain “ijk”.
- matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.
- num_range("x", 1:3): matches x1, x2 and x3.

See ?select for more details.

Use the rename() function to rename variables that aren't explicitly mentioned (would be dropped if renaming with the select() function.)
```{r}
rename(flights, tail_num = tailnum) # tailnum -> tail_num; added the underscore
```

Another option is to use select() in conjunction with the everything() helper. This is useful if you have a handful of variables you’d like to move to the start of the data frame.
```{r}
select(flights, time_hour, air_time, everything())
```

### mutate()
Used to add new columns (to th end of the dataset) that are functions of existing columns.

```{r}
# create a narrower flights dataset to more easily view all columns
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)

# add new columns with mutate()
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

```{r}
# refer to the newly created columns
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

Use transmute to only keep the new variables
```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

### summarise()
Collapses a data frame into a single row

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
# all aggregation function have an na-rm arg to remove missing values
```

summarise() is not terribly useful unless we pair it with group_by(). This changes the unit of analysis from the complete dataset to individual groups. The output is a grouped summary, one of the most commonly used tools when working with dplyr.

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

#### the pipe

Explore the relationship between distance and average delay for each location
```{r}
by_dest <- group_by(flights, dest)  # group flights by destination
delay <- summarise(by_dest, # summarize the compute distance, avg delay, & # of flights
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL") # filter to remove noise & Honolulu

# It looks like delays increase with distance up to ~750 miles 
# and then decrease. Maybe as flights get longer there's more 
# ability to make up delays in the air?
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

Can achieve the same 3 steps using the pipe, %>%
```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

#### counts
Good idea to include a count or count of non-missing values to ensure you're not drawing conclusions from small amounts of data.

Example: planes with the highest avg delays
```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```

Draw a scatterplot of the above
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

Filter out the groups with the smallest numbers of observations (often the largest sources of variation)
```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) + 
    geom_point(alpha = 1/10)
```

#### grouping by multiple variables
When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll up a dataset:
```{r}
daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))
```

```{r}
(per_month <- summarise(per_day, flights = sum(flights)))
```

```{r}
(per_year  <- summarise(per_month, flights = sum(flights)))
```

#### ungrouping
Use the ungroup() function to remove grouping or return to operations on ungrouped data
```{r}
daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights
```

#### grouped mutates (and filters)
A grouped filter is a grouped mutate followed by an ungrouped filter

Find the worst members of each group:
```{r}
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```

Find all groups bigger than a threshold:
```{r}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
popular_dests
```

Standardise to compute per group metrics:
```{r}
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)
```

## Exploratory Data Analysis
To do data cleaning, you’ll need to deploy all the tools of EDA: visualisation, transformation, and modelling

Load packages
```{r}
library(tidyverse)
```

There is no rule about which questions you should ask to guide your research. However, two types of questions will always be useful for making discoveries within your data. You can loosely word these questions as:
- What type of variation occurs within my variables?
- What type of covariation occurs between my variables?

### Variation
Variation is the tendency of the values of a variable to change from measurement to measurement.

Use a bar chart to examine the distribution of a categorical variable
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

The height of the bars displays how many observations occurred with each x value. You can compute these values manually with dplyr::count():
```{r}
diamonds %>% 
  count(cut)
```

Use a histogram to examine the distribution of a continuous variable
```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

You can compute this by hand by combining dplyr::count() and ggplot2::cut_width():
```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

Explore a different binwidth
```{r}
smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)
```

Overlay multiple histograms in the same plot with geom_freqpoly() instead of using the geom_histogram arg - uses lines instead of bars
```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

### Missing Values
Options when dealing with unusual values:

Drop the entire row with strange values
```{r}
diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))
```

Recommended: replace the unusual values with missing values
```{r}
diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```

ggplot won't plot missing values, but will provide a warning when they've been removed
```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()
```

Compare observations w/ and w/o missing values by making a new variable with is.na()
```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time), # missing values here
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

### Covariation
Covariation is the tendency for the values of two or more variables to vary together in a related way. The best way to spot covariation is to visualise the relationship between two or more variables.

#### A categorical and continuous variable
Ex: explore how the price of a diamond varies with its quality
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

```{r}
ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

Display density (counts standardized such that the AUC of each frequency polygon = 1) instead of counts
```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

Look at price distribution using a boxplot:
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

Use the reorder() to order categorical variables for better visualization
```{r}
# default view
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
```

```{r}
# reorder class based on the median value of hwy
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))
```

Transpose the boxplot for better viewing of long variable names
```{r}
# use coord_flip() to transpose a boxplot (flip by 90 degrees)
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()
```

#### Two categorical variables
Count the number of observations for each combination using geom_count():
```{r}
ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))
```

Can also compute the count with dplyr:
```{r}
diamonds %>% 
  count(color, cut)
```

Then visualize with geom_tile() and the fill aesthetic:
```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
```

#### Two continuous variables
Draw a scatterplot with geom_point():
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

Scatterplots are less useful with larger datasets b/c points start to overplot.
Fix this issue by adding transparency with the alpha aesthetic:
```{r}
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 100)
```

Use geom_bin2d() and geom_hex() to divide a coordinate plane into 2D bins:
```{r}
ggplot(data = smaller) +
  geom_bin2d(mapping = aes(x = carat, y = price)) # rectangular bins

# install.packages("hexbin")
ggplot(data = smaller) +
  geom_hex(mapping = aes(x = carat, y = price)) # hexagonal bins
```

Can also bin one continuous variable so that it behaves like a categorical variable
```{r}
# bin carat, then display a boxplot for each group
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))
```

Use cut_number() to display approximately the same number of points in each bin
```{r}
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))
```

### Patterns and models
Patterns reveal co-variation, so that we know to account for them in our analysis.

Example: It’s hard to understand the relationship between cut and price, because cut and carat, and carat and price are tightly related. It’s possible to use a model to remove the very strong relationship between price and carat so we can explore the subtleties that remain. The following code fits a model that predicts price from carat and then computes the residuals (the difference between the predicted value and the actual value). The residuals give us a view of the price of the diamond, once the effect of carat has been removed:
```{r}
library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))
```

Removing the co-variate/confounding variable "carat" then allows one to properly visualize the relationship between cut and price:
```{r}
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))
# relative to their size, better quality diamonds are more expensive
```

